---
title: "3.1"
format: html
---

{{< include activity-instructions.qmd >}}


```{r}
#| eval: true

# Import data
hikes <- read.csv("https://mac-stat.github.io/data/high_peaks.csv")
```
### Exercise 1: Research Questions {.unnumbered}

Let's dig into the `hikes` data, starting with the `elevation` and difficulty `ratings` of the hikes:

```{r}
#| eval: true

head(hikes)
```

a.  What features would we like a visualization of the *categorical* difficulty `rating` variable to capture?

b.  What about a visualization of the *quantitative* `elevation` variable

# COMMENT on the change in the code and the corresponding change in the plot
ggplot(hikes, aes(x = rating)) +
  geom_bar()
# COMMENT on the change in the code and the corresponding change in the plot
ggplot(hikes, aes(x = rating)) +
  geom_bar() +
  labs(x = "Rating", y = "Number of hikes")
# COMMENT on the change in the code and the corresponding change in the plot
ggplot(hikes, aes(x = rating)) +
  geom_bar(color = "orange", fill = "blue") +
  labs(x = "Rating", y = "Number of hikes")
# COMMENT on the change in the code and the corresponding change in the plot
ggplot(hikes, aes(x = rating)) +
  geom_bar(color = "orange", fill = "blue") +
  labs(x = "Rating", y = "Number of hikes")
# COMMENT on the change in the code and the corresponding change in the plot
ggplot(hikes, aes(x = rating)) +
  geom_bar(color = "orange", fill = "blue")  +
  labs(x = "Rating", y = "Number of hikes") +
  theme_minimal()
  
